---
title: "DelayedDataFrame: an on-disk represention of DataFrame"
author: 
- name: Qian Liu
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Hervé Pagès
  affiliation: Fred Hutchinson Cancer Research Center, Seattle, WA
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "last edit: 07/22/2018"
output:
    BiocStyle::html_document:
        toc: true
        toc_float: true
package: DelayedDataFrame
vignette: |
    %\VignetteIndexEntry{DelayedDataFrame: an on-disk representation of data in DataFrame structure}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

```{r options, eval=TRUE, echo=FALSE}
options(showHeadLines=3)
options(showTailLines=3)
```

# Introduction

As the genetic/genomic data are having increasingly larger profile,
the annotation file are also getting much bigger than expected. the
memory space in R has been an obstable for fast and efficient data
processing, because most available _R_ or _Bioconductor_ packages are
developed based on in-memory data manipulation. With some newly
developed data structure as [HDF5][] or [GDS][], and the R interface of
[DelayedArray][] to represent on-disk data structures with different
back-end in R-user-friendly array data structure (e.g.,
[HDF5Array][],[GDSArray][]), the high-throughput genetic/genomic data
are now being able to easily loaded and manipulated within R. However,
the annotation files for the samples and features inside the
high-through data are also getting unexpectedly larger than
before. With an ordinary `data.frame` or `DataFrame`, it is still
getting more and more challenging for any analysis to be done within
R. So here we have developed the `DelayedDataFrame`, which has the
very similar characteristics as `data.frame` and `DataFrame`. But at
the same time, all column data could be optionally saved on-disk
(e.g., in [DelayedArray][] structure with any back-end). Common
operations like constructing, subsetting, splitting, combining could
be done in the same way as `DataFrame`. This feature of
`DelayedDataFrame` could enable efficient on-disk reading and
processing of the large-scale annotation files, and at the same,
signicantly saves memory space with common `DataFrame` metaphor in
_R_ and _Bioconductor_.

[HDF5]: https://www.hdfgroup.org/solutions/hdf5/
[GDS]: http://corearray.sourceforge.net/
[DelayedArray]: https://bioconductor.org/packages/DelayedArray
[GDSArray]: https://bioconductor.org/packages/GDSArray
[HDF5Array]: https://bioconductor.org/packages/HDF5Array


# Installation
Download the package from Bioconductor: 

```{r getPackage, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("DelayedDataFrame")
```

The development version is also available to download through github: 

```{r getDevel, eval=FALSE}
devtools::install_github("Bioconductor/DelayedDataFrame")
```
Load the package into R session before using:

```{r Load, message=FALSE, warning=FALSE}
library(DelayedDataFrame)
```

# DelayedDataFrame class

## class extension

`DelayedDataFrame` extends the `DataFrame` data structure, with an additional slot called `lazyIndex`, which saves all the mapping indexes for each column of the data inside `DelayedDataFrame`. It is similar to `data.frame` in terms of construction, subsetting, splitting, combining... The `rownames` are having same feature as `DataFrame`. It will not be given automatically, but only by explicitly specify in the constructor function `DelayedDataFrame(, row.names=...)` or using the slot setter function `rownames()<-`. 

```{r, construction}
da1 <- DelayedArray(array(1:10))
da2 <- DelayedArray(array(11:30, dim=c(10,2)))
(ddf <- DelayedDataFrame(da0=letters[1:10], da1=I(da1), da2=I(da2)) )
```

## slot accessors

The slots of `DelayedDataFrame` could be accessed by `lazyIndex()`, `nrow()`, `rownames()` (if not NULL) functions. With a newly constructed `DelayedDataFrame` object, the initial value of `lazyIndex` slot will be NULL for all columns. 

```{r, accessors}
lazyIndex(ddf)
nrow(ddf)
rownames(ddf)
```

## `lazyIndex` slot

The `lazyIndex` slot is in `LazyIndex` class, which is defined in the
`DelayedDataFrame` package and extends the `SimpleList` class. The
`listData` slot saves unique indexes for all the columns, and the
`index` slots saves the position of index in `listData` slot for each
column in `DelayedDataFrame` object.  In the above example, with an
initial construction of `DelayedDataFrame` object, the index for each
column will all be NULL, and all 3 columns points the NULL values
which sits in the first position in `listData` slot of `lazyIndex`.

```{r}
lazyIndex(ddf)@listData
lazyIndex(ddf)@index
```

Whenever an operation is done (e.g., subsetting), the `listData` slot
inside the `DelayedDataFrame` stays the same, but the `lazyIndex` slot
will be updated, so that the show method, further statistical
calculation will be applied to the subsetting data set.  For example,
here we subset the `DelayedDataFrame` object `ddf` to keep only the
first 5 rows, and see how the `lazyIndex` works. As shown in below,
after subsetting, the `listData` slot in `ddf1` stays the same as
`ddf`. But the subsetting operation was recorded in the `lazyIndex`
slot, and the slots of `lazyIndex`, `nrows` and `rownames` (if not
NULL) are all updated. So the subsetting operation is kind of
`delayed`.

```{r, lazyIndex}
ddf1 <- ddf[1:5,]
identical(ddf@listData, ddf1@listData)
lazyIndex(ddf1)
nrow(ddf1)
```

Only when functions like `DataFrame()`, or `as.list()`, the 
`lazyIndex` will be realized and `DelayedDataFrame` returned. 
We will show the realization in the following coercion method section.     


# DelayedDataFrame methods

The common methods on `data.frame` or `DataFrame` are also defined on
`DelayedDataFrame` class, so that they behave similarily on
`DelayedDataFrame` objects.

## Coercion methods

Coercion methods between `DelayedDataFrame` and other data structures
are defined. When coercing from `ANY` to `DelayedDataFrame`, the
`lazyIndex` slot will be added automatically, with the initial NULL
value of indexes for each column.

- From vector

```{r}
as(letters, "DelayedDataFrame")
```

- From DataFrame

```{r}
as(DataFrame(letters), "DelayedDataFrame")
```

- From list
```{r}
(a <- as(list(a=1:5, b=6:10), "DelayedDataFrame"))
lazyIndex(a)
```

When coerce `DelayedDataFrame` into other data structure, the
`lazyIndex` slot will be realized and the new data structure
returned. For example, when `DelayedDataFrame` is coerced into a
`DataFrame` object, the `listData` slot will be updated according to
the `lazyIndex` slot.


```{r}
df1 <- as(ddf1, "DataFrame")
df1@listData
dim(df1)
```

## Subsetting methods

### subsetting by `[`

two-dimensional `[` subsetting on `DelayedDataFrame` objects by
integer, character, logical values all work.

- integer subscripts. 

```{r, singleSB1}
ddf[, 1, drop=FALSE]
```

- character subscripts (column names).

```{r, singleSB2}
ddf[, "da1", drop=FALSE]
```

- logical subscripts. 

```{r, singleSB3}
ddf[, c(TRUE,FALSE,TRUE)]
```

When subsetting using `[` on an already subsetted `DelayedDataFrame`
object, the `lazyIndex`, `nrows` and `rownames`(if not NULL) slot will
be updated.

```{r, singleSB4}
a <- ddf1[1:4, 2, drop=FALSE]
lazyIndex(a)
nrow(a)
```

### subsetting by `[[`

The `[[` subsetting will take column subscripts for integer or
character values, and return corresponding columns in it's original
data format.

```{r, doubleSB}
ddf[[2]]
ddf[["da2"]]
```

## `rbind/cbind`

When doing `rbind`, the `lazyIndex` of input arguments will be
realized and a new `DelayedDataFrame` with NULL lazyIndex will be
returned.

```{r, rbind}
ddf2 <- ddf[6:10, ]
(ddfrb <- rbind(ddf1, ddf2))
lazyIndex(ddfrb)
```

`cbind` of `DelayedDataFrame` objects will keep all existing
`lazyIndex` of input arguments and carry into the new
`DelayedDataFrame` object.

```{r, cbind, error=FALSE}
(ddfcb <- cbind(da0=ddf1[,1], ddf1[, 2:3]))
## lazyIndex(ddfcb)
```

# sessionInfo

```{r, sessioninfo}
sessionInfo()
```
