--- 
title: "DelayedDataFrame"
subtitle: on-disk data representation in `DataFrame` metaphor
author: 
- name: Qian Liu
  affiliation: Bioconductor Core Team, RPCCC, Buffalo
date: "`r Sys.Date()`"
output: ioslides_presentation
---

## DelayedDataFrame class

- Extends: DataFrame 
```{r, DataFrameClass}
library(DelayedDataFrame)
(ddf <- DelayedDataFrame())
slotNames(ddf)
```

- `lazyIndex` slot  
Saves the mapping indexes for each column.

## Constructor: 

Each argument in `...` is coerced to a `DataFrame` and combined column-wise.
```
DelayedDataFrame(..., row.names=FALSE, check.names=TRUE)
```

```{r, constructor}
obj <- DelayedDataFrame(letters, LETTERS, row.names=LETTERS)
lazyIndex(obj)
```

## On-disk data representation in `DataFrame` format

```{r, GDSArray}
library(GDSArray)
file <- SeqArray::seqExampleFileName("gds")
gdsnodes(file)
```

---

Use `GDSArray` to represent the GDS nodes for variant annotation.

```{r, GDSArray1}
varid <- GDSArray(file, "annotation/id")
AA <- GDSArray(file, "annotation/info/AA")
```

```{r, GDSArrayStr, tidy=TRUE}
seed(varid)
```

---

Construct a `DelayedDataFrame` object with `GDSArray` columns.

```{r, ddfgds}
(ddf <- DelayedDataFrame(varid, AA))
```

## subsetting

- `listData`: unchanged

```{r, subset}
(ddf1 <- ddf[1:20,])
identical(ddf@listData, ddf1@listData)
```

--- 

- `lazyIndex`: updated
- `nrows`: updated
- `rownames`: updated (if not NULL)

```{r, subset1}
lazyIndex(ddf1)
nrow(ddf1)
```

## `lazyIndex` realization

```{r, coercion}
as(ddf1, "DataFrame")
```

